\chapter{Discussion \& Conclusion}\label{chap:discussion-conclusion}
\section{Considerations}

The experiments show that the proposed correlation metrics do capture structure in the data.
This is most visible in the controlled test workload, where the planner can separate favorable and unfavorable query--filter pairs and use this signal to improve the speed--accuracy trade-off.
In particular, the adaptive planner reaches essentially the same recall as ACORN while obtaining higher throughput.

At the same time, the test workload should be interpreted carefully, since the metadata attributes used were constructed to have a clear geometric relationship to the embedding space.
This makes it a useful environment for validating whether the metrics can detect the intended signal, but it also makes the setting close to ideal.
If a metric fails or becomes unstable in this setting, this is therefore a stronger warning sign than it would be on a noisy external dataset.

\paragraph{Metric conservatism at low selectivity.}

This concern is especially relevant when interpreting the \(c_{\mathrm{tanh}}\) and \(c_{\mathrm{pgap}}\) metrics.
Both metrics detect structure in the controlled experiments, but they also begin to classify some queries as random even in this favorable setting.
This corresponds to the collapse toward zero discussed in Chapter~\ref{chap:experiments}.

In practice, this conservatism is acceptable.
At very low selectivities, PRE-FILTER is usually the best strategy regardless of the correlation signal, because the filtered candidate set is already small.
In this range, selectivity alone is often sufficient to choose a good plan.

Correlation-aware decisions become more important at higher selectivities, where PRE-FILTER becomes more expensive and the choice between POST-FILTER and ACORN is less obvious.
These are also the cases where local and global predicate distributions are easier to distinguish, since \(p_c\) and \(p_g\) are less likely to vanish because of sparse counts.
Thus, even if \(c_{\mathrm{tanh}}\) and \(c_{\mathrm{pgap}}\) are conservative for very narrow predicates, they remain useful in the selectivity range where correlation-aware planning matters most.

\paragraph{SIFT and GIST as random-attribute baseline.}

The SIFT and GIST validation results need particular attention.
As described in Section~\ref{subsec:sift-and-gist}, the SIEVE attributes used for these datasets are themselves synthetic: the numeric attribute columns are sampled independently from a standard normal distribution and then assigned to the base vectors.
They are therefore not real semantic metadata attributes, and they are not constructed to have a meaningful relationship to the embedding geometry.

For both datasets, \(c_{\mathrm{tanh}}\) and \(c_{\mathrm{pgap}}\) classify all queries as random.
Given the way the attributes are generated, this is not necessarily a failure; for independent attributes, a correlation metric should mostly report no structure.
\(c_{\mathrm{heur}}\) is more active on SIFT, where it classifies \(1{,}498\) out of \(10{,}000\) queries as non-random: \(932\) negative and \(566\) positive.
It also classifies all GIST queries as random.

To understand whether the SIFT cases flagged by \(c_{\mathrm{heur}}\) correspond to actual metadata--embedding structure, 
we compared the \(1{,}498\) non-random heuristic labels against the true correlation metric already defined in Section~\ref{sec:correlation},
\[
    c_{\mathrm{true}}
    =
    \frac{\mathbb{E}[g_k(q,R_p)] - g_k(q,X_p)}
         {\mathbb{E}[g_k(q,R_p)]},
\]

\begin{table}[H]
    \centering
    \renewcommand{\arraystretch}{1.12}
    \begin{tabular}{lrrr}
        \toprule
        \(c_{\mathrm{heur}}\) label & True negative & True random & True positive \\
        \midrule
        Negative & 442 & 431 & 59 \\
        Positive & 30 & 360 & 176 \\
        \bottomrule
    \end{tabular}
    \caption{Distance-based true-correlation labels for SIFT queries classified as non-random by \(c_{\mathrm{heur}}\).}
    \label{tab:sift-heuristic-true-correlation}
\end{table}

The result is mixed.
Among the SIFT queries flagged as non-random by \(c_{\mathrm{heur}}\), only \(41.3\%\) agree with the distance-based sign, \(52.8\%\) are actually random.
As a fraction of the full SIFT workload, this means that from the \(14.9\%\) non-random queries flagged by \(c_{\mathrm{heur}}\) only \(6.18\%\) are confirmed by \(c_{\mathrm{true}}\).
Thus the heuristic signal on SIFT is real enough to produce non-random labels, but it is noisy and should not be interpreted as robust evidence of metadata--embedding correlation.

\paragraph{Type I and Type II errors.}

This also raises a more general question about the fallibility of the metric.
Any lightweight runtime proxy can make two kinds of mistakes.
A Type I error occurs when the metric classifies a query as positively or negatively correlated even though the query behaves like a random one.
A Type II error occurs when the metric classifies a query as random even though a true positive or negative correlation exists.
For query planning, Type I errors are more harmful.
They can route a query to a specialized execution strategy based on a correlation signal that is not actually present, producing unnecessary recall loss or latency overhead.
Type II errors are still undesirable, because the planner may miss an opportunity to exploit structure, but the fallback behavior is usually closer to a safe baseline.

In this thesis, we therefore prefer a conservative metric that avoids false correlation claims over a more active metric that detects more cases but produces many unreliable non-random labels.

\paragraph{App-review validation.}

To further analyse the behaviour observed in Section~\ref{subsec:app-reviews}, we compare each proposed metric against \(c_{\mathrm{true}}\).
For both \texttt{text\_length} and \texttt{star\_rating}, we audit three groups of queries: those with the strongest positive scores, those with the strongest negative scores, and those closest to zero.
From each group, we sample 25 queries and compute \(c_{\mathrm{true}}\) for each of them.
The corresponding label is obtained using the calibrated threshold \(\tau(s)\), following the procedure described in Section~\ref{sec:tau}.

Table~\ref{tab:app-review-validation} reports the agreement between each proposed metric and \(c_{\mathrm{true}}\).
Spearman correlation measures whether the two scores induce a similar ranking of the audited queries, while Pearson correlation measures the strength of their linear relationship.

\begin{table}[H]
    \centering
    \renewcommand{\arraystretch}{1.12}
    \begin{tabular}{llrr}
        \toprule
        Attribute & Metric & Spearman & Pearson \\
        \midrule
        Star rating & Heuristic & 0.8998 & 0.7824 \\
        Star rating & Pgap      & 0.8794 & 0.7680 \\
        Star rating & Tanh      & 0.8521 & 0.7866 \\
        \midrule
        Text length & Heuristic & 0.5682 & 0.4016 \\
        Text length & Pgap      & 0.7270 & 0.4080 \\
        Text length & Tanh      & 0.6760 & 0.5299 \\
        \bottomrule
    \end{tabular}
    \caption{Agreement between metric scores and \(c_{\mathrm{true}}\) on app-review queries.}
    \label{tab:app-review-validation}
\end{table}

All three metrics are positively correlated with \(c_{\mathrm{true}}\) for both attributes.
This means that the cheaper metric scores generally follow the more expensive \(c_{\mathrm{true}}\) computation.
The agreement is strongest for \texttt{star\_rating}, where all metrics show high Spearman and Pearson correlation.

For \texttt{text\_length}, the signal is weaker, and the differences between the metrics become clearer.
The heuristic and pgap metrics still preserve the ranking reasonably well, but their Pearson correlation drops to around \(0.40\), meaning that their scores do not follow the scale of \(c_{\mathrm{true}}\) as closely.
Tanh is more stable: it keeps a higher Pearson correlation for \texttt{text\_length}, while showing similar behaviour to the other metrics on \texttt{star\_rating}.
This makes it the most robust metric in this validation.

\paragraph{Metric signal versus planner quality.}

Furthermore, the app-review results also show that detecting structure is not sufficient by itself.
Even when the metrics identify a correlation signal, the adaptive planner does not consistently outperform the fixed baselines.
POST-FILTER remains the fastest strategy, PRE-FILTER remains the highest-recall strategy, and ACORN usually provides the strongest fixed speed--recall trade-off.
The planner is sometimes faster than ACORN, but it often loses too much recall to dominate it.
This is especially visible for \texttt{star\_rating}, where many queries are routed to POST-FILTER and the planner remains well below ACORN in recall.
For \texttt{text\_length}, the routing is somewhat more balanced and the planner recovers more recall, but the improvement is still not enough to clearly beat ACORN.

This raises two fundamental questions.

The first question is whether the metric can detect structure in the data.
The experiments suggest that it can, especially in the controlled workload and in the app-review validation set.

The second question is whether the current routing policy uses that structure optimally.
The validation results suggest that it does not.
The planner uses a simple rule-based policy derived from the synthetic experiments, and that policy is too conservative in some real-world cases.

\paragraph{Recommended metric.}

Based on these observations, we recommend \(c_{\mathrm{tanh}}\) as the default correlation metric.
It retains the same conceptual structure as the heuristic formulation, but downweights weak lift values smoothly.
This avoids turning very small differences between \(p_c\) and \(p_g\) into confident positive or negative scores.
This makes it more reliable than \(c_{\mathrm{heur}}\), whose SIFT labels agree with the distance-based true signal for only \(41.3\%\) of flagged cases.

We also prefer \(c_{\mathrm{tanh}}\) over \(c_{\mathrm{pgap}}\).
Both metrics are conservative on the independent-attribute SIFT and GIST workloads, where they classify all queries as random.
However, \(c_{\mathrm{pgap}}\) depends on the absolute probability difference \(p_c-p_g\), which can suppress rare predicates even when they are meaningfully enriched in the local neighborhood.
In practice, the two metrics behave similarly, but \(c_{\mathrm{tanh}}\) appears more robust in the app-review validation: it remains competitive on \texttt{star\_rating} and achieves a higher Pearson correlation on \texttt{text\_length}.

\section{Future work}

\paragraph{Validation in database systems.}

A natural next step is to evaluate the proposed metrics inside a real database system.
The experiments in this thesis isolate the filtered vector search problem in order to make the behavior of the different strategies easier to analyze. 
They therefore do not capture the full complexity of a production query engine, where components such as query optimization, index maintenance, buffer management and concurrent workloads can also influence search performance.

In the current implementation, statistics such as cluster assignments, local attribute distributions, global distributions, and cumulative distribution functions are built explicitly for the experimental pipeline.
Inside a database system, these statistics would instead need to become part of the system metadata and be made available to the query planner during optimization.

A particularly important question is which parts of the metric can be computed offline.
For example, the JS divergence between each cluster distribution and the global distribution could be materialized in advance, since it does not depend on the query predicate.
Similarly, the per-cluster CDFs could be stored so that, at runtime, the planner only needs to assign the query vector to a cluster and compute the probability mass of the filter interval.
This would make the metric much cheaper to evaluate during query planning.

Future work should therefore study the engineering cost of maintaining these statistics in a real system.
This includes measuring how long it takes to build the per-cluster histograms and CDFs, how this cost scales with the number of vectors and attributes, and how often the statistics would need to be refreshed as the database changes.
The number of clusters is also an important design parameter: more clusters may capture local correlation more accurately, while fewer clusters may be cheaper and more stable.
Although this introduces a time--space trade-off, the main question is whether these statistics can be maintained cheaply enough for the metric to be useful in a production query optimizer.

\paragraph{Parameter tuning.}

A second direction is to move beyond the discrete classification used in this thesis.
Here, the planner reduces the estimated correlation to one of three regimes: positive, negative, or random.
In practice, two positively correlated queries may differ substantially in how strongly the admissible predicate subset is concentrated near the query neighborhood.
Likewise, two negatively correlated queries may require very different amounts of additional search effort.

Future work should therefore treat correlation as a continuous signal rather than only as a discrete routing label.
The goal would be to estimate how much a filter predicate changes the expected difficulty of the search, and to use this estimate to dynamically tune the internal parameters of filtered vector search methods.

For example, in ACORN~\cite{acorn}, the correlation score could guide the amount of search effort.
A locally enriched predicate may require a lower \texttt{efSearch} to reach a target recall, since many nearby candidates are likely to satisfy the filter.
A locally depleted predicate, on the other hand, may require a higher \texttt{efSearch}, because the search must explore further before finding enough admissible candidates.

This follows a similar intuition to DARTH~\cite{darth}, which uses a learned model to decide when an ANN search can terminate while still meeting a target recall.
The proposed correlation score could be added as an input feature to such a recall prediction model.
The model would then combine search-state features, such as distance computations and candidate-distance statistics, with a query-specific estimate of predicate--embedding correlation, 
and could make overall dynamic recall prediction more robust.

Rather than selecting a completely different execution strategy at hard thresholds, the system could continuously adjust search parameters to match the estimated workload difficulty.
Such a formulation may also reduce the instability caused by boundary cases: queries close to a threshold would not abruptly switch plans, but would instead receive slightly different search budgets.
More broadly, this direction could allow correlation-aware methods to complement existing state-of-the-art techniques rather than replace them, effectively improving ACORN, DARTH, or future filtered ANN systems.