\chapter{Correlation Metric}
Our metric captures two aspects simultaneously: (i) how different the query-local attribute distribution is from the global dataset distribution, and (ii) whether the queried filter range is locally enriched or depleted.

Formally, let \(X = \{x_1,\dots,x_N\} \subseteq \mathbb{R}^d\) be the dataset of embedding vectors, and let each vector \(x_i\) be associated with a scalar attribute \(a_i\). 
Given a query vector \(q \in \mathbb{R}^d\) and a filter range \([L,U]\), we seek to determine whether values in \([L,U]\) are more likely in the neighborhood of \(q\) than in the dataset as a whole.

\section{Precomputed statistics}
We precompute three components: 

\subsection{Cluster centroids.}
Let \(\ell_i \in \{1,\dots,k\}\) denote the cluster assignment of vector \(x_i\). 
For each cluster \(c\), define
\[
C_c = \{ i \mid \ell_i = c \}.
\]
The centroid is reconstructed as
\[
c_c = \frac{1}{|C_c|} \sum_{i \in C_c} x_i.
\]

\subsection{Attribute distributions.}
We estimate both global and cluster-local attribute distributions using histogram-based density estimation over the scalar attribute \(a_i\).

\textbf{Bin construction.}
The number of histogram bins $M$ is determined adaptively from the data.
We begin by estimating the bin width using the Freedman--Diaconis rule,
\[
h = \frac{2 \cdot \mathrm{IQR}}{n^{1/3}},
\]
where $\mathrm{IQR}$ is the interquartile range of the attribute values and $n$
is the number of samples. Given the data range $[v_{\min}, v_{\max}]$, this
yields the bin-count estimate
\[
\widehat{M} = \left\lceil \frac{v_{\max} - v_{\min}}{h} \right\rceil.
\]
When the interquartile range is too small (i.e., the data is nearly constant),
this estimate becomes unreliable, and we instead fall back to a heuristic based
on a target number of samples per bin.

To keep the number of bins reasonable, we force it to stay within a fixed range:
\[
M = \min\!\bigl(M_{\max},\, \max(M_{\min},\, \widehat{M})\bigr).
\]
This way, the histogram is neither too rough (losing detail) nor too detailed
(with many empty bins that make the results unreliable).

In our implementation, we fix $M_{\min} = 20$ and $M_{\max} = 200$. Since the
dataset characteristics are known, these bounds are
sufficient and yield stable results in practice.

More generally, these bounds can be defined in a data-dependent manner. For
example, letting $u$ denote the number of distinct attribute values, one may use
\[
M_{\min} = \min\!\left(u,\, \max\!\left(5, \left\lfloor n^{1/3} \right\rfloor \right)\right),
\qquad
M_{\max} = \min\!\left(u,\, \left\lfloor \frac{n}{m_{\min}} \right\rfloor \right),
\]
where $m_{\min}$ is the minimum desired number of samples per bin. Such choices
adapt the binning resolution to both dataset size and attribute cardinality.

\textbf{Global distribution.}
Given bin edges \(b_0 < b_1 < \dots < b_M\), we first compute a histogram over
the full dataset:
\[
H_{\mathrm{global}}(j)
=
\left|
\{\, i \mid a_i \in [b_j, b_{j+1}) \,\}
\right|,
\qquad j=0,\dots,M-1.
\]

We then convert this histogram into a probability distribution. To avoid zero
probabilities, we apply additive smoothing by adding a small constant
\(\varepsilon > 0\) to each bin:
\[
P_{\mathrm{global}}(j)
=
\frac{H_{\mathrm{global}}(j) + \varepsilon}
{\sum_{k=0}^{M-1} \bigl(H_{\mathrm{global}}(k) + \varepsilon\bigr)}.
\]
This ensures that no bin has zero probability.
Without smoothing, bins with zero counts would lead to undefined or unstable
values in downstream computations (e.g., divisions or logarithms in divergence
measures).

After smoothing, the histogram no longer sums to one, so we renormalize it.
We simply rescale all values so that they form a valid probability
distribution again:
\[
\sum_j P_{\mathrm{global}}(j) = 1.
\]

The corresponding cumulative distribution function is
\[
\mathrm{CDF}_{\mathrm{global}}(j)
=
\sum_{k=0}^{j} P_{\mathrm{global}}(k).
\]

\textbf{Cluster-local distributions.}
For each cluster \(c\), we first compute its radius
\[
r_c = \max_{i \in C_c} \|x_i - c_c\|_2.
\]
We then introduce a \textbf{scaling parameter} \(\lambda \ge 1\) to slightly enlarge the
cluster and define the set of selected points
\[
C'_c = \{\, i \in C_c \mid \|x_i - c_c\|_2 \le \lambda r_c \,\}.
\]
This allows us to include points near the boundary of the cluster, making the
estimate more robust. The choice and effect of \(\lambda\) are discussed in
Section~\ref{subsec:radius-expansion}.

Using these points, we estimate the cluster-level attribute distribution by
constructing a histogram over the same bins:
\[
H_c(j)
=
\left|
\{\, i \in C'_c \mid a_i \in [b_j, b_{j+1}) \,\}
\right|,
\qquad j=0,\dots,M-1,
\]
which corresponds to the empirical distribution
\[
P_c(a) = \frac{1}{|C'_c|} \sum_{i \in C'_c} \mathbf{1}[a_i \in \text{bin}(a)].
\]
We then convert it into a probability distribution by applying additive
smoothing and renormalization, as in the global case:
\[
P_c(j)
=
\frac{H_c(j) + \varepsilon}
{\sum_{k=0}^{M-1} \bigl(H_c(k) + \varepsilon\bigr)}.
\]
Finally, the cumulative distribution function is computed analogously:
\[
\mathrm{CDF}_c(j)
=
\sum_{k=0}^{j} P_c(k).
\]

\paragraph{Query-conditional distribution.}
We consider the \emph{query-conditional attribute distribution}
\(P(a \mid x)\), which describes the distribution of attribute values in the
neighborhood of a point \(x \in \mathbb{R}^d\). Intuitively, it captures how
attribute values are distributed among vectors that are similar to \(x\) in the
embedding space.

Estimating \(P(a \mid x)\) directly is challenging, as it requires defining a
local neighborhood around each query and computing a distribution from a
potentially small number of samples. Instead, we approximate the mapping
\[
x \mapsto P(a \mid x)
\]
by assigning a single distribution \(P_c(a)\) to each cluster:
\[
x \mapsto P_c(a), \quad \text{for } x \in R_c,
\]
where \(R_c\) denotes the region associated with cluster \(c\). This avoids
estimating a separate distribution for every query location, which would be
both computationally expensive and statistically unreliable.

\paragraph{Distributions and estimation.}
We distinguish between three distributions. 

The conditional distribution
\(P(a \mid x)\) is defined at the point level. 

At the cluster level, we
introduce an ideal (population-level) distribution \(P_c(a)\), which we define
as the average of the conditional distributions over the region \(R_c\):
\[
P_c(a) = \mathbb{E}_{x' \sim \mathrm{Unif}(R_c)}[P(a \mid x')].
\]
This provides a clean abstraction of the attribute distribution associated
with the cluster.

In practice, this distribution is not directly observable and is instead
estimated empirically using a histogram over the points in the cluster:
\[
\hat{P}_c(a) = \frac{1}{n_c} \sum_{i \in C'_c} \mathbf{1}[a_i \in \text{bin}(a)].
\]

\subsection{Query-to-cluster mapping.}
Given a query \(q \in \mathbb{R}^d\), we associate it with a representative
region of the dataset by assigning it to its nearest centroid. To this end, we
construct an index over the set of centroids \(\{c_1, \dots, c_k\}\) and define
\[
c^*(q) = \arg\min_c \|q - c_c\|_2.
\]
This induces a partition of the vector space into Voronoi regions and assigns
each query to the cluster whose centroid is closest to it.

To estimate the attribute distribution in the vicinity of \(q\), we use the
empirical distribution of its assigned cluster, given by the histogram-based
estimate computed in the previous section. 
\[
P(a \mid q) \approx P_{c^*(q)}(a),
\]
where \(P_{c^*(q)}\) denotes the attribute distribution associated with the
cluster \(c^*(q)\). 

\section{Approximation quality}\label{sec:approximation-quality}
We now analyze how well the cluster-based approximation
\[
P(a \mid q) \approx P_c(a)
\]
captures the true query-conditional distribution.

\paragraph{Smoothness assumption}
We assume that attribute distributions are correlated with the underlying
vector space geometry, meaning that nearby points exhibit similar attribute
behavior. 

Formally, we assume a \emph{Lipschitz continuity} condition: there exists a
constant \(L > 0\) such that
\[
D(P(a \mid x), P(a \mid x')) \le L \|x - x'\|
\]
for all \(x, x' \in \mathbb{R}^d\), where \(D(\cdot,\cdot)\) measures the
difference between two distributions.

This means that the change in the attribute distribution is proportional to the
distance between points: small movements in the embedding space lead to small
changes in the distribution, while large differences can only occur between
points that are far apart.

\subsection{Bounding the approximation error}

For a point \(x \in R_c\), the approximation error is
\[
D(P(a \mid x), P_c(a)),
\]
where \(D(\cdot,\cdot)\) is a divergence between distributions.

Recall that we define the ideal cluster-level distribution as the average of
the conditional distributions over the region \(R_c\), i.e., \(P_c(a)\) is the
average of \(\{P(a \mid x') : x' \in R_c\}\).

We now bound the deviation between \(P(a \mid x)\) and this average. Since
\(P_c(a)\) is an average over these distributions, its distance to
\(P(a \mid x)\) can be bounded by the largest deviation within the cluster:
\[
D(P(a \mid x), P_c(a))
\le
\sup_{x' \in R_c} D(P(a \mid x), P(a \mid x')).
\]

Applying the smoothness assumption gives:
\[
D(P(a \mid x), P_c(a))
\le
\sup_{x' \in R_c} D(P(a \mid x), P(a \mid x'))
\]
\[
D(P(a \mid x), P_c(a))
\le
\sup_{x' \in R_c} L \|x - x'\|
\]
\[
D(P(a \mid x), P_c(a))
\le
L \sup_{x' \in R_c} \|x - x'\|.
\]

Taking the supremum over all \(x \in R_c\) yields:
\[
\sup_{x \in R_c} D(P(a \mid x), P_c(a))
\le
L \sup_{x,x' \in R_c} \|x - x'\|.
\]

\paragraph{Cluster diameter.}
We define the diameter of a cluster \(R_c\) as
\[
\mathrm{diam}(R_c) = \sup_{x,x' \in R_c} \|x - x'\|.
\]
Substituting this into the previous bound yields
\[
\sup_{x \in R_c} D(P(a \mid x), P_c(a))
\le
L \, \mathrm{diam}(R_c).
\]


\subsection{Error decomposition}

The proposed approximation introduces two main sources of error. 

The first is the quantization (approximation) error,
\[
P(a \mid x) \;\;\text{vs}\;\; P_c(a),
\]
which arises from replacing the pointwise conditional distribution with a
cluster-level approximation. 

The second is the estimation error,
\[
P_c(a) \;\;\text{vs}\;\; \hat{P}_c(a),
\]
which is introduced when estimating the cluster-level distribution from a
finite number of samples using a histogram.

\subsubsection{Quantization error (Geometric Discretization).}
For a query assigned to cluster \(c\), we define the worst-case approximation
error as
\[
\varepsilon_{\mathrm{quant}}(q)
=
\sup_{x \in R_c} D(P(a \mid x), P_c(a)),
\]
where \(D(\cdot,\cdot)\) is a divergence between distributions (e.g., total
variation or Jensen--Shannon). This measures the error introduced by replacing
the pointwise distribution \(P(a \mid x)\) with a single cluster-level
distribution.

From the previous bound, this error satisfies
\[
\varepsilon_{\mathrm{quant}}(q)
\le
L \, \mathrm{diam}(R_c),
\]
where \(\mathrm{diam}(R_c)\) is the cluster diameter and \(L\) is the smoothness
constant. Thus, the approximation error increases with the size of the cluster
and with how rapidly the distribution changes in the embedding space.

However, this bound is worst-case and does not fully reflect practical behavior.
In particular, the error also depends on how the geometry of the cluster aligns
with the structure of \(P(a \mid x)\).

\paragraph{Refinements beyond diameter.}
Three factors are particularly important:

\emph{Cluster anisotropy.}
The diameter only captures the maximum pairwise distance, but ignores the shape
of the cluster. As a result, clusters can be anisotropic (e.g., elongated or
skewed).

Even if most points are close to the centroid, an elongated cluster can still
contain points that are far apart. Along the short axis, points are very
similar, while along the long axis, points may differ significantly. Thus, two
points in the same cluster can have very different attribute distributions
despite being assigned the same \(P_c(a)\).

\emph{Alignment with distribution gradients.}
Think of \(P(a \mid x)\) as assigning, to every point \(x\) in the embedding
space, a full distribution over attribute values:
\[
x \mapsto P(a \mid x).
\]
If we fix a location \(x\) and look at vectors near \(x\), the
attribute values of those vectors follow a distribution \(P(a \mid x)\). Moving
to a different point \(x'\) may change this distribution. Thus, as we move
through the embedding space, we can think of the attribute distribution as
evolving over space.

\textit{Case 1: Good alignment (low error).}
If a cluster lies within a region where this distribution changes little, then
for any two points \(x, x' \in R_c\), we have \(P(a \mid x) \approx P(a \mid x')\).
In this case, all points in the cluster exhibit similar attribute behavior, and
using a single distribution \(P_c(a)\) is a good approximation.

\textit{Case 2: Poor alignment (high error).}
If a cluster spans a region where the distribution changes significantly, then
points within the cluster may have very different attribute distributions.
However, the model still assigns a single \(P_c(a)\), which forces these
different behaviors to be averaged together and introduces systematic bias.

For example, suppose that in one region of the space the attributes are
concentrated around 10, while in another region they are concentrated around
100. If a cluster spans both regions, then \(P_c(a)\) mixes these two distinct
distributions and fails to accurately represent either.

\emph{Local curvature.}
Finally, the smoothness assumption only captures first-order (linear) variation.
If \(P(a \mid x)\) varies nonlinearly, the Lipschitz bound may underestimate the
true variation within the cluster. As a result, even clusters with small
diameter can incur non-negligible approximation error.

\subsubsection{Estimation error (Statistical Estimation).}
Recall the ideal (population-level) definition of the cluster distribution:
\[
P_c(a) = \mathbb{E}_{x' \in R_c}[P(a \mid x')].
\]
This represents the true cluster-level distribution, which is not directly
observable.

In practice, we estimate this distribution empirically using a histogram over
the points in the cluster:
\[
\hat{P}_c(a) = \frac{1}{n_c} \sum_{i \in C'_c} \mathbf{1}[a_i \in \text{bin}(a)],
\]
where \(n_c = |C'_c|\) is the number of samples in the cluster. This estimator
relies on discretizing the attribute values into bins.

The estimation error is defined as
\[
\varepsilon_{\mathrm{est}}(c)
=
D(\hat{P}_c(a), P_c(a)),
\]
which measures the deviation between the empirical estimate and the true
cluster-level distribution. This error arises from several sources:

\emph{Finite sampling noise (variance).}
We estimate the distribution using a finite number of samples, which introduces
sampling noise that decreases with the sample size:
\[
\varepsilon_{\mathrm{est}}(c) \sim \mathcal{O}\!\left(\frac{1}{\sqrt{n_c}}\right).
\]
This scaling corresponds to the standard deviation of empirical estimates. Thus, large clusters yield more
stable estimates, while small clusters produce noisy distributions.

\emph{Histogram discretization (bias).}
We do not estimate the full distribution, but a binned version
\(P_c(a) \approx P_c(\text{bin}(a))\).
As a result, fine structure within bins is lost and values near bin boundaries
may be distorted. This introduces systematic bias, as the resolution of the
distribution is reduced.

\emph{Interaction with cluster size.}
Small clusters have few samples (\(n_c\)), leading to high variance and unstable
estimates of \(P_c(a)\). Large clusters have more samples and therefore lower
variance, but tend to cover more heterogeneous regions, which increases the
approximation error.

\subsection{Bias--variance trade-off}

The proposed method introduces a trade-off between approximation
(quantization) error and estimation error. This trade-off is controlled by two
parameters: the number of clusters \(k\) and the radius expansion parameter
\(\lambda\).

\subsubsection{Effect of the number of clusters \(k\).}

The number of clusters \(k\) determines the granularity of the partition of the
embedding space.

Increasing \(k\) reduces the cluster diameter, leading to lower quantization
error, but also reduces the number of samples per cluster, resulting in higher
estimation error.

Decreasing \(k\) increases the cluster diameter, leading to higher
quantization error, but increases the number of samples per cluster, resulting
in lower estimation error.

\subsubsection{Effect of radius expansion \(\lambda\).}\label{subsec:radius-expansion}

In addition to clustering, we can expand each cluster region using a parameter
\(\lambda > 1\). We define the expanded region
\[
r'_c = \lambda \, r_c, \quad
R'_c = \{ x : \|x - c_c\| \le r'_c \}.
\]

The diameter of the expanded region satisfies
\[
\mathrm{diam}(R'_c) = \sup_{x,x' \in R'_c} \|x - x'\|.
\]
Using the triangle inequality,
\[
\|x - x'\| \le \|x - c_c\| + \|x' - c_c\| \le r'_c + r'_c = 2r'_c,
\]
which implies
\[
\mathrm{diam}(R'_c) \le 2 r'_c = 2 \lambda r_c.
\]

Thus, \(\lambda\) directly controls an upper bound on the cluster diameter, and
consequently the quantization error.

Increasing \(\lambda\) enlarges the effective region \(R'_c\), increasing the
diameter (and thus the potential quantization error), while also increasing the
number of samples \(n_c\), which reduces estimation variance. Decreasing
\(\lambda\) has the opposite effect: it reduces the diameter and quantization
error, but also reduces the sample size, increasing estimation variance.

Overall, \(\lambda\) acts as a regularization parameter, trading off geometric
fidelity against statistical stability by controlling the spatial extent over
which the conditional distribution is assumed constant.

\section{Correlation Metric}
We define the correlation metric that quantifies the alignment between a query and a filter. 

\subsection{Alignment Scoring.}
Given a query $q$, an associated cluster $c^* = c^*(q)$ and a filter defined by an attribute range $[L,U]$, we quantify the alignment between the filter and the local attribute distribution.

Using the cumulative distribution functions, we compute the probability mass of the filter range under both the local and global distributions:
\[
P_{c^*}([L,U]) = \mathrm{CDF}_{c^*}(U) - \mathrm{CDF}_{c^*}(L),
\]
\[
P_{\mathrm{GLOBAL}}([L,U]) = \mathrm{CDF}_{\mathrm{GLOBAL}}(U) - \mathrm{CDF}_{\mathrm{GLOBAL}}(L).
\]
These quantities measure how likely it is, under each distribution, for a randomly drawn attribute value to fall within the filter interval.

\subsection{Lift.}
We define the lift as the logarithmic ratio between local and global probabilities:
\[
\mathrm{lift}(q) = \log \left( \frac{P_{c^*}([L,U]) + \epsilon}{P_{\mathrm{GLOBAL}}([L,U]) + \epsilon} \right),
\]
where $\epsilon > 0$ is a small constant for numerical stability.

The lift measures the relative enrichment or depletion of the filter range within the local distribution compared to the global baseline.

\paragraph{Relative vs.\ Absolute Comparison.}
We use a ratio rather than a difference because we are interested in relative deviations. An absolute difference 
\[
P_{c^*}([L,U]) - P_{\mathrm{GLOBAL}}([L,U])
\]
can be misleading, as it does not account for the scale of the probabilities.

Example: $0.02$ vs $0.01$ $\rightarrow$ difference $= 0.01$ (small), but actually doubled.\\
Example: $0.6$ vs $0.5$ $\rightarrow$ difference $= 0.1$ (bigger), but only $20\%$ increase.

By contrast, the ratio
\[
\frac{P_{c^*}([L,U])}{P_{\mathrm{GLOBAL}}([L,U])}
\]
directly captures how many times more (or less) likely the filter range is under the local distribution compared to the global distribution. Values greater than one indicate enrichment, while values less than one indicate depletion.

If $\mathrm{lift}(q) > 0$, the filter range contains more probability mass locally than globally.\\
If $\mathrm{lift}(q) < 0$, the filter range is underrepresented locally.

Thus, $\mathrm{lift}$ encodes the direction of alignment between the filter and the local distribution.

\paragraph{Logarithmic Transformation.}
We apply a logarithm to the ratio for two reasons.

First, it yields a symmetric representation around $0$. Without the logarithm, enrichment and depletion are not symmetric:
\[
2 \quad \text{vs} \quad 0.5.
\]
With the logarithm:
\[
\log(2) = -\log(0.5).
\]
Thus:
\[
\text{Enriched: } \log(2) \approx 0.69, \quad
\text{Depleted: } \log(0.5) \approx -0.69.
\]

Second, ratios can explode. For example:
\[
P_{\mathrm{GLOBAL}} = 0.001, \quad P_{c^*} = 0.1 \;\Rightarrow\; \text{ratio} = 100.
\]
The logarithm compresses this:
\[
\log(100) \approx 4.6,
\]
preventing extreme values from dominating.

\subsection{Divergence.}
To quantify the extent to which the local attribute distribution differs from the global distribution, we compute the Jensen--Shannon divergence:
\[
\mathrm{JS}(P_{c^*} \,\|\, P_{\mathrm{GLOBAL}}).
\]

The Jensen--Shannon divergence is defined as
\[
\mathrm{JS}(P \,\|\, Q) = \tfrac{1}{2} \mathrm{KL}(P \,\|\, M) + \tfrac{1}{2} \mathrm{KL}(Q \,\|\, M),
\quad M = \tfrac{1}{2}(P + Q),
\]
where $\mathrm{KL}$ denotes the Kullback--Leibler divergence:
\[
\mathrm{KL}(P \,\|\, Q) = \sum_j P(j) \log \frac{P(j)}{Q(j)}.
\]

The Kullback--Leibler divergence measures how inefficient it would be to represent samples drawn from $P$ using a distribution $Q$, or equivalently, how much information is lost when $Q$ is used in place of $P$. However, it is directional.

The Jensen--Shannon divergence converts this into a symmetric measure of dissimilarity by comparing both distributions to their midpoint $M$. It therefore captures how different the two distributions are without privileging either one.

In our setting, $\mathrm{JS}(P_{c^*} \,\|\, P_{\mathrm{GLOBAL}})$ measures how strongly the local attribute distribution associated with the query differs from the overall dataset-wide distribution. Small values indicate similarity, while larger values indicate distinct structure.

\paragraph{Why Not KL Divergence Directly?}
Although KL divergence is standard, it is not ideal here.

First, it is asymmetric:
\[
\mathrm{KL}(P \,\|\, Q) \neq \mathrm{KL}(Q \,\|\, P).
\]

Second, it can be undefined when $Q(j) = 0$ and $P(j) > 0$.

Third, it is unbounded above, making it harder to interpret.

By contrast, Jensen--Shannon divergence is symmetric, always defined, and bounded:
\[
0 \le \mathrm{JS}(P \,\|\, Q) \le \log 2.
\]

\subsection{Final Scores.}\label{subsec:final-scores}

We consider multiple variants of the correlation score to study how alignment direction and magnitude interact.

\paragraph{Signed Interpretation (Baseline).}
For ${C_{\mathrm{heur}}}$, the score decomposes as:
\[
C_{\mathrm{heur}} = \mathrm{JS}(P_{c^*} \,\|\, P_{\mathrm{GLOBAL}}) \cdot \mathrm{sign}(\mathrm{lift}(q)).
\]

The divergence measures how strongly the local distribution deviates from the global baseline, while the lift determines whether the filter range is locally enriched or depleted.

If $\mathrm{JS}(P_{c^*} \,\|\, P_{\mathrm{GLOBAL}}) \approx 0$, then $\mathrm{corr}_1(q) \approx 0$, regardless of the lift.

A positive score indicates enrichment,
$P_{c^*}([L,U]) > P_{\mathrm{GLOBAL}}([L,U])$, while a negative score indicates depletion,
$P_{c^*}([L,U]) < P_{\mathrm{GLOBAL}}([L,U])$.

Thus, $C_{\mathrm{heur}}$ captures both the strength of distributional deviation and the direction of alignment.

\paragraph{Tanh-scaled Interpretation.}
For $C_{\tanh}$, the score is given by:
\[
C_{\tanh} = \mathrm{JS}(P_{c^*} \,\|\, P_{\mathrm{GLOBAL}}) \cdot \tanh(\mathrm{lift}(q)).
\]

The divergence again measures the deviation between local and global distributions, but the lift now influences both the direction and the magnitude of the score through a smooth, bounded transformation.

As in the baseline, the sign of the score indicates enrichment or depletion. However, stronger alignment leads to larger magnitudes, while extreme values are dampened by the $\tanh$ function.

If $\mathrm{JS}(P_{c^*} \,\|\, P_{\mathrm{GLOBAL}}) \approx 0$, then $C_{\tanh} \approx 0$, regardless of the lift.

Thus, $C_{\tanh}$ reflects both distributional deviation and the strength of alignment in a continuous manner.

\paragraph{Difference-based Interpretation.}
For $C_{\mathrm{pgap}}$, the score is defined as:
\[
C_{\mathrm{pgap}} = \mathrm{JS}(P_{c^*} \,\|\, P_{\mathrm{GLOBAL}}) \cdot \big(P_{c^*}([L,U]) - P_{\mathrm{GLOBAL}}([L,U])\big).
\]

The divergence again captures the deviation between local and global distributions, while the difference term measures the absolute change in probability mass.

The sign indicates enrichment or depletion, as in the previous formulations. However, the magnitude depends on absolute differences rather than relative ones, making the score sensitive to the scale of the probabilities.

If $\mathrm{JS}(P_{c^*} \,\|\, P_{\mathrm{GLOBAL}}) \approx 0$, then $C_{\mathrm{pgap}} \approx 0$, regardless of the difference term.

Thus, $C_{\mathrm{pgap}}$ combines distributional deviation with absolute, rather than relative, alignment effects.


\section{Thresholding with \texorpdfstring{$\tau$}{tau}}\label{sec:tau}

The correlation score is a continuous quantity. However, query planning requires a discrete classification of the correlation type: \emph{negative}, \emph{random}, or \emph{positive}. To obtain this, we introduce a symmetric threshold $\tau \ge 0$ and define
\[
\hat{y}(s;\tau)=
\begin{cases}
\text{positive}, & s>\tau,\\[4pt]
\text{negative}, & s<-\tau,\\[4pt]
\text{random}, & |s|\le \tau.
\end{cases}
\]
Thus, $\tau$ determines the minimum magnitude a score must have to be considered informative, while values close to zero are treated as uncorrelated.

\subsection{Ground Truth.}
We use the true correlation $C(q,p)$ (Section~\ref{sec:correlation}) as ground truth. Each query is assigned an intended class (\emph{negative}, \emph{random}, or \emph{positive}) based on the sign and magnitude of $C(q,p)$. The goal is to choose $\tau$ such that the predicted labels $\hat{y}(s;\tau)$, derived from an approximate score $s$, best match these ground truth classes.

\subsection{Evaluation metrics.}
For a given metric, we evaluate candidate thresholds
\[
\tau \in [0, \max_i |s_i|],
\]
where $\{s_i\}$ are the observed scores. This interval is discretized using a uniform grid (400 points by default). For each candidate $\tau$, all queries are classified according to the rule above, yielding predicted labels.

We then compare predicted and intended labels using the following metrics:
\[
\text{accuracy}, \quad \text{balanced accuracy}, \quad \text{macro-F1}.
\]

Let $y_i$ and $\hat{y}_i$ denote the true and predicted labels for query $i$, and let $\mathcal{C} = \{\text{negative}, \text{random}, \text{positive}\}$.

\paragraph{Accuracy.}
Accuracy measures overall agreement between predicted and true labels:
\[
\mathrm{Accuracy} = \frac{1}{N} \sum_{i=1}^{N} \mathbf{1}\{\hat{y}_i = y_i\},
\]
where $N$ is the number of queries and $\mathbf{1}\{\cdot\}$ is the indicator function.

\paragraph{Balanced Accuracy.}
Balanced accuracy averages recall across all classes:
\[
\mathrm{BalancedAcc} = \frac{1}{|\mathcal{C}|} \sum_{c \in \mathcal{C}} \mathrm{Recall}_c,
\]
where
\[
\mathrm{Recall}_c = \frac{\mathrm{TP}_c}{\mathrm{TP}_c + \mathrm{FN}_c},
\]
and $\mathrm{TP}_c$ and $\mathrm{FN}_c$ denote the number of true positives and false negatives for class $c$.

\paragraph{Macro-F1.}
Macro-F1 averages the F1 score across classes:
\[
\mathrm{MacroF1} = \frac{1}{|\mathcal{C}|} \sum_{c \in \mathcal{C}} \mathrm{F1}_c,
\]
where
\[
\mathrm{F1}_c = \frac{2 \cdot \mathrm{Precision}_c \cdot \mathrm{Recall}_c}{\mathrm{Precision}_c + \mathrm{Recall}_c},
\]
and
\[
\mathrm{Precision}_c = \frac{\mathrm{TP}_c}{\mathrm{TP}_c + \mathrm{FP}_c}.
\]

Accuracy measures overall agreement but can be misleading under class imbalance (e.g., when most queries are classified as \emph{random}). Balanced accuracy and macro-F1 instead give equal weight to each class, making them more appropriate for evaluating threshold quality in this setting.

\subsection{Selected Threshold.}
The selected threshold is
\[
\tau^* = \arg\max_{\tau} \mathrm{Obj}(\tau),
\]
where $\mathrm{Obj}$ is the chosen evaluation metric (balanced accuracy by default). If multiple thresholds achieve the same optimum, the smallest one is selected. 

\subsection{Selectivity Dependence.}\label{subsec:selectivity-dependence}
The optimal threshold varies with the selectivity $s$ of the filter. To account for this, we estimate $\tau$ at multiple selectivity levels and fit a function
\[
\tau(s) \approx p(\log_{10} s),
\]
where $p$ is a polynomial. This provides a smooth approximation of the threshold across different selectivity regimes.
This is dealt with more detail in Section~\ref{sec:tau-estimation}.

\subsection{Usage in Query Planning.}
The function $\tau(s)$ is stored as a database statistic. At query time, the planner estimates the selectivity $s$, computes $\tau(s)$, and classifies the correlation score accordingly. The resulting class (\emph{negative}, \emph{random}, or \emph{positive}) determines the query execution strategy.