\chapter{Introduction}
This thesis explores how global IPv6 addressing increases the attack surface of the Internet of Things (IoT) environment. The study combines not only a comprehensive literature review but also an empirical assessment of default system configurations through controlled laboratory demonstrations, highlighting how the adoption of new network standards can potentially affect the security of millions of end IoT devices.
% to polish

\section{The IPv4 to IPv6 Transition}
is the result of an inherent limitation in the design of IPv4’s address space, which only supports $2^{32}$ unique addresses. A boundary already exceeded by the estimated 16 billion IoT devices alone in 2024~\cite{iotanalytics2024}. IPv6 mitigates this limitation by implementing a 128-bit architecture, offering a virtually infinite address space. Despite its technical advantage, the global adoption has been gradual due to the current scale of IPv4-based infrastructures and the cost of migration. Therefore, technologies such as Network Address Translation (NAT) were adopted to extend the lifetime of IPv4~\cite{rfc3022}.

In practice, NAT allows multiple internal hosts within a private local-area network (LAN) to share a single public IPv4 address by translating internal IP addresses and TCP/UDP port numbers to a public address and port at the network boundary. This translation process is stateful and connection-oriented, meaning that the address mappings are created only for outbound traffic initiated by hosts within the LAN~\cite{rfc3022,rfc4787}. Consequently, unless explicitly allowed through static port-forwarding rules, any unsolicited inbound traffic that does not match an existing mapping entry is discarded by default ~\cite{rfc4787}. Nonetheless, the use of NAT translation tables violates the original end-to-end connectivity principle of the Internet architecture~\cite{saltzer1984end}, and simultaneously provides a form of perimeter protection by preventing direct external access to internal hosts. 

\section{Loss of NAT and Increased IoT Exposure}
reflects a shift in network architecture paradigms. Until now, many IoT devices have been designed under an implicit threat model that assumes limited inbound reachability, relying primarily on network topology rather than on end-device security mechanisms. With the deprecation of NAT under IPv6, this assumption no longer holds. Devices become directly reachable without any change to their internal security properties, thereby increasing their effective exposure~\cite{saltzer1984end,rfc8200}.
The direct exposure of IoT devices is especially problematic due to the attack vectors commonly associated with this type of devices. The study \emph{Understanding the Mirai Botnet}~\cite{antonakakis2017mirai} demonstrates that a large number of IoT devices have exposed services which are only protected by weak or default authentication mechanisms, enabling automated large-scale exploitation. Similarly, \emph{A Large-Scale Analysis of the Security of Embedded Firmware}~\cite{costin2014firmware} reveals that embedded devices frequently contain insecure network services or misconfigurations, as well as persistent vulnerabilities produced by design constraints and limited update mechanisms. Moreover, \emph{Security Analysis of Emerging Smart Home Applications}~\cite{fernandes2016smarthome} shows how poorly secured management interfaces and API calls significantly expand the attack surface when IoT devices become externally reachable. It is precisely the interaction between inherent device vulnerabilities and expanded reachability that increases the risk of remote exploitation.
\section{Research Objectives \& Questions}
The main objective of this thesis is to analyze how the transition to globally routable IPv6 addressing increases the attack surface of IoT devices by enabling direct external exposure in real-world network environments. Specifically, this study aims to identify under which router configurations, network parameters, and default device settings IoT devices become directly accessible from external networks. 

\section{Contributions}
%TODO

\section{Thesis Structure}
This thesis follows the next structure. Chapter~2 provides the necessary technical background on IPv4, IPv6, modems configurations and IoT security models. Chapter~3 describes the experimental methodology and demo setup used to evaluate device exposure under different network configurations. Chapter~4 presents the empirical results and analyzes the conditions under which IoT devices become externally reachable. Chapter~5 discusses the security implications of these findings and outlines practical considerations for reducing unintended exposure. Finally, Chapter~6 concludes the thesis and summarizes key insights and possible future developments for the research.
% to polish
