\begin{acknowledgements}
I would like to sincerely thank Prof. Dr. Yannis Chronis and Dr. Vasilis Mageirakos for their valuable input, guidance, and support throughout this thesis. I am especially grateful for the freedom and trust they gave me to explore the topic independently, while still providing helpful direction whenever needed.

Mulțumesc mult, Mama, Bunica și Cruci, pentru că m-ați susținut mereu, pentru încrederea voastră constantă și pentru că ați crezut mereu în mine. Sprijinul vostru a fost mereu unul dintre lucrurile care mi-au dat putere să merg mai departe.

Grazie ai miei fratelli Lucien, Penno e Kevin, che per me sono una seconda casa e che mi riportano a terra quando ce n'è bisogno. Grazie per esserci sempre stati, per le risate, per i consigli e per farmi sentire a casa ovunque io sia.

Muchas gracias, Miguel, por la mano que nos hemos dado en estos años, cubriéndonos las espaldas el uno al otro. Gracias por estar siempre ahí, por compartir tantos momentos importantes y por hacer que este percorso fuera más ligero.

Thank you, Andrea and Philipp, for being there regardless of the situation, and for lifting me back up whenever I needed it. I am grateful for your patience, your presence, and the way you have supported me through both the easy and difficult parts of my journey.

Thank you, Felix and Luis, for everything I have learned from you, for broadening my perspective, and for listening to me rant about this project. Our conversations helped me think differently, stay grounded, and see this work from angles I would not have considered on my own.

Finally, I want to thank myself, without whom none of this work would have been done.
\end{acknowledgements}

